RILS-ROLS: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-10-22 doc/paper.tex
Small update.
Perturbations pseudo-code.

    \begin{algorithm}
   	\begin{algorithmic}[1] 
   		\State \textbf{Input}: input training dataset $D_{tr}$
   		\State \textbf{Output}: best symbolic formula solution $bs$
   		\State $n_{tr} \gets |D_{tr}|$
   		\State $sample_{size} \gets $ \texttt{InitialSampleSize}($n_{tr}$)
   		\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)
   		\State $s \gets NodeConstant(0)$ 
   		\State $s_{fit} \gets$ \texttt{Fitness}($s, D_{tr}'$)
   		\State $bs, bs_{fit} \gets s, s_{fit}$ \label{line:solSet}
   		\State $start_{tried}, perturbations_{tried} \gets \emptyset, \emptyset$
   		\While{\emph{stopping criteria is not met}}
   			\State $start_{tried} \gets start_{tried} \cup \{s\}$
   			\State $s_{perturbations} \gets $ \texttt{Generate1Perturbations}($s$) \label{line:sPert}
   			\State $s_{perturbations} \gets $ \texttt{FitOLS}($s_{perturbations}$, $D_{tr}'$)
   			\State $s_{perturbations} \gets $ \texttt{OrderByR2}($s_{perturbations}$) \label{line:orderR2}
   			\State $improved \gets false $
   			\For{$p \in s_{perturbations}$}
   				\If{$p \in perturbations_{tried}$}
   					\State \textbf{continue}
   				\EndIf
  				\State $p \gets $ \texttt{Simplify}($p$) \label{line:simp}
  				\State $p \gets $ \texttt{LocalSearch}($p$, $D_{tr}'$) \label{line:ls}
  				\State $p_{fit} \gets$ \texttt{Fitness}($p$, $D_{tr}'$)
  				\If{$p_{fit} < bs_{fit}$} \label{line:avoid1}
  					\State $bs, bs_{fit}, improved \gets p, p_{fit}, true$ // new best solution
  					\State \textbf{break}
  				\EndIf \label{line:avoid2}
  				\State $perturbations_{tried} \gets perturbations_{tried} \cup \{p\}$
   			\EndFor
   			\If{improved}
   				\State $s \gets bs$ \label{line:improved}
   			\Else
   				\State $start_{candidates} \gets $ \texttt{Generate1Perturbations}($bs$)
   				\If{$start_{candidates} \setminus start_{tried} = \emptyset$} // all 1-perturbations around $bs$ tried
   					\State $s' \gets $ \texttt{RandomPick}($start_{candidates}$)  \label{line:pert21}
   					\State $start_{candidates} \gets $ \texttt{Generate1Perturbations}($s'$) \label{line:pert22}
   				\EndIf
   				\State $s \gets $ \texttt{RandomPick}($start_{candidates} \setminus start_{tried}$) \label{line:randPick}
   				\If{not improved for too many iterations}
   					\State $sample_{size} \gets$ \texttt{IncreaseSampleSize}($sample_{size}, n_{tr}$) \label{line:sampleAdj1}
   					\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)\label{line:sampleAdj2}
   				\EndIf
   			\EndIf
   			\If{$R^2$ almost 1 and RMSE almost 0}
   				\State \textbf{break} // early exit
   			\EndIf
   		\EndWhile
   		\State \Return $bs$
   	\end{algorithmic}
   	\caption{RILS-ROLS algorithm.}
   	\label{alg:rilsrols}
   \end{algorithm}

@@

On of the first steps in the main loop is to generate perturbations near the starting solution $s$ (line~\ref{line:sPert}). 
As the name of this procedure (\texttt{Generate1Perturbations}) suggests, perturbation is local -- meaning the closeness of starting solution $s$ and any of perturbations is 1 (we call it 1-perturbation sometimes). The precise way of generating perturbation is described in a separate Section~\ref{sec:pertGen}.

@@

If some of $s_{perturbations}$ around starting solution $s$ yielded an improvement, $bs$ becomes the starting solution $s$ in the next iteration of the main loop (line~\ref{line:improved}). 
Otherwise, it makes no sense to set starting solution to $bs$, since nothing changed -- search is \emph{trapped} in local optimum. Therefore, a randomness needs to be introduced. First, a set of local perturbations around $bs$ is generated ($start_{candidates}$) in the same manner as before (procedure \texttt{Generate1Perturbations}). If at least one of these was not previously used as a starting solution ($start_{tried}$), a single perturbation from the $start_{candidates} \setminus start_{tried}$ is randomly picked (line~\ref{line:randPick}). There is small chance that $start_{candidates} \setminus start_{tried} = \emptyset$. When that happens, the set of starting solution candidates is equal to perturbations of some randomly selected perturbation of $bs$ (lines \ref{line:pert21}-\ref{line:pert22}) -- which effectively means that perturbations on distance 2 from $bs$ are used. There is a very small chance that these 2-perturbations will have empty set difference with $start_{tried}$ set of starting solutions. Therefore, 3-perturbations are not considered.

@@

@@

>>>>

    \begin{algorithm}
    \small
    \hspace*{\algorithmicindent} \textbf{Input}: input training dataset $D_{tr}$ \\
    \hspace*{\algorithmicindent} \textbf{Output}: best symbolic formula solution $bs$
   	\begin{algorithmic}[1] 
   		\Procedure{RILS-ROLS}{$D_{tr}$}
   		\State $n_{tr} \gets |D_{tr}|$
   		\State $sample_{size} \gets $ \texttt{InitialSampleSize}($n_{tr}$)
   		\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)
   		\State $s \gets NodeConstant(0)$ 
   		\State $s_{fit} \gets$ \texttt{Fitness}($s, D_{tr}'$)
   		\State $bs, bs_{fit} \gets s, s_{fit}$ \label{line:solSet}
   		\State $start_{tried}, perturbations_{tried} \gets \emptyset, \emptyset$
   		\While{\emph{stopping criteria is not met}}
   			\State $start_{tried} \gets start_{tried} \cup \{s\}$
   			\State $s_{perturbations} \gets $ \texttt{All1Perturbations}($s$) \label{line:sPert}
   			\State $s_{perturbations} \gets $ \texttt{FitOLS}($s_{perturbations}$, $D_{tr}'$)
   			\State $s_{perturbations} \gets $ \texttt{OrderByR2}($s_{perturbations}$) \label{line:orderR2}
   			\State $improved \gets false $
   			\For{$p \in s_{perturbations}$}
   				\If{$p \in perturbations_{tried}$}
   					\State \textbf{continue}
   				\EndIf
  				\State $p \gets $ \texttt{Simplify}($p$) \label{line:simp}
  				\State $p \gets $ \texttt{LocalSearch}($p$, $D_{tr}'$) \label{line:ls}
  				\State $p_{fit} \gets$ \texttt{Fitness}($p$, $D_{tr}'$)
  				\If{$p_{fit} < bs_{fit}$} \label{line:avoid1}
  					\State $bs, bs_{fit}, improved \gets p, p_{fit}, true$ // new best solution
  					\State \textbf{break}
  				\EndIf \label{line:avoid2}
  				\State $perturbations_{tried} \gets perturbations_{tried} \cup \{p\}$
   			\EndFor
   			\If{improved}
   				\State $s \gets bs$ \label{line:improved}
   			\Else
   				\State $start_{candidates} \gets $ \texttt{All1Perturbations}($bs$)
   				\If{$start_{candidates} \setminus start_{tried} = \emptyset$} // all 1-perturbations around $bs$ tried
   					\State $s' \gets $ \texttt{RandomPick}($start_{candidates}$)  \label{line:pert21}
   					\State $start_{candidates} \gets $ \texttt{Generate1Perturbations}($s'$) \label{line:pert22}
   				\EndIf
   				\State $s \gets $ \texttt{RandomPick}($start_{candidates} \setminus start_{tried}$) \label{line:randPick}
   				\If{not improved for too many iterations}
   					\State $sample_{size} \gets$ \texttt{IncreaseSampleSize}($sample_{size}, n_{tr}$) \label{line:sampleAdj1}
   					\State $D_{tr}' \gets$ \texttt{Sample}($D_{tr}, sample_{size}$)\label{line:sampleAdj2}
   				\EndIf
   			\EndIf
   			\If{$R^2$ almost 1 and RMSE almost 0}
   				\State \textbf{break} // early exit
   			\EndIf
   		\EndWhile
   		\State \Return $bs$
   		\EndProcedure
   	\end{algorithmic}
   	\caption{RILS-ROLS method.}
   	\label{alg:rilsrols}
   \end{algorithm}

@@

On of the first steps in the main loop is to generate perturbations near the starting solution $s$ (line~\ref{line:sPert}). 
As the name of this procedure (\texttt{All1Perturbations}) suggests, perturbation is local -- meaning the closeness of starting solution $s$ and any of perturbations is 1 (we call it 1-perturbation sometimes). The precise way of generating perturbation is described in a separate Section~\ref{sec:pertGen}.

@@

If some of $s_{perturbations}$ around starting solution $s$ yielded an improvement, $bs$ becomes the starting solution $s$ in the next iteration of the main loop (line~\ref{line:improved}). 
Otherwise, it makes no sense to set starting solution to $bs$, since nothing changed -- search is \emph{trapped} in local optimum. Therefore, a randomness needs to be introduced. First, a set of local perturbations around $bs$ is generated ($start_{candidates}$) in the same manner as before (procedure \texttt{All1Perturbations}). If at least one of these was not previously used as a starting solution ($start_{tried}$), a single perturbation from the $start_{candidates} \setminus start_{tried}$ is randomly picked (line~\ref{line:randPick}). There is small chance that $start_{candidates} \setminus start_{tried} = \emptyset$. When that happens, the set of starting solution candidates is equal to perturbations of some randomly selected perturbation of $bs$ (lines \ref{line:pert21}-\ref{line:pert22}) -- which effectively means that perturbations on distance 2 from $bs$ are used. There is a very small chance that these 2-perturbations will have empty set difference with $start_{tried}$ set of starting solutions. Therefore, 3-perturbations are not considered.

@@

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: solution $s$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local perturbations (1-perturbations) around solution $s$ -- $s_{perturbations}$
	\begin{algorithmic}[1] 
		\Procedure{All1Perturbations}{$s$}
		\State $s_{perturbations} \gets \emptyset$
		\State $s \gets$ \texttt{NormalizeConstants}($s$)
		\State $s \gets$ \texttt{Simplify}($s$)
		\State $s_{subtrees} \gets$ \texttt{SubTrees}($s$)
		\For{$ss \in s_{subtrees}$}
			\If{$ss = s$}
				\State $s_{perturbations} \gets s_{perturbations} \cup$ \texttt{Node1Perturbations}($ss$)
			\EndIf
			\If{$ss.arity \geq 1$}
				\State $node_{perturbations} \gets$ \texttt{Node1Perturbations}($ss.left$)
				\For{$np \in node_{perturbations}$}
					\State $ss_{new} \gets$ \texttt{Replace}($ss$, $ss.left$, $np$)
					\State $s_{perturbations} \gets s_{perturbations} \cup \{ss_{new}\}$
				\EndFor
			\EndIf
			\If{$ss.arity = 2$}
				\State $node_{perturbations} \gets$ \texttt{Node1Perturbations}($ss.right$)
				\For{$np \in node_{perturbations}$}
					\State $ss_{new} \gets$ \texttt{Replace}($ss$, $ss.right$, $np$)
					\State $s_{perturbations} \gets s_{perturbations} \cup \{ss_{new}\}$
				\EndFor
			\EndIf
		\EndFor
		\State \Return $s_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of 1-perturbations.}
	\label{alg:pert}
\end{algorithm}

@@

\subsection{Datasets and SRBench}
